---
title: "HW14: alpha diversity"
author: "Bonnie Hurwitz"
date: "`r Sys.Date()`"
output:
  word_document:
    toc: yes
    toc_depth: '3'
  pdf_document:
    toc: yes
    toc_depth: '3'
  html_document:
    theme: paper
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: yes
subtitle: Exploring Alpha Diversity
editor_options:
  markdown:
    wrap: sentence
---

# Setup

```{r setup, include=FALSE}
# Change your NETID
knitr::opts_chunk$set(echo = TRUE, fig.width = 6, fig.height = 4, dpi = 120)
knitr::opts_knit$set(root.dir = '/xdisk/bhurwitz/bh_class/NETID/assignments/14_alpha_diversity')
```

```{r}
getwd()
```

```{r message=FALSE}
library(seriation)
library(dplyr)
library(purrr)
library(ggplot2)
library(phyloseq)
library(microViz)
library(shiny)
```

## Importing and fixing taxa your FMT dataset

To get started, we need to import our example.biom file into R and work with it using the phyloseq package

```{r}
# replace your NETID and SETID here, for example set21.biom
fmt <- import_biom("/xdisk/bhurwitz/bh_class/NETID/assignments/12_phyloseq/SETID.biom")
```

### Fixing tax_table

Let's format the dataset.

```{r}
fmt@tax_table@.Data <- substring(fmt@tax_table@.Data, 4)
colnames(fmt@tax_table@.Data)<- c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")
fmt <- tax_fix(fmt, verbose = FALSE)
```

### Starting here

Copy your notebook into the current working directory. Be sure to change to your NETID.

Also, take a minute to "knit" the Rmd notebook into word, and submit this on D2L. Note that this will become useful later for your report, when you select a few graphs to include.

```{bash, engine.opts='-l'}
# copy your notebook into the working directory
cp ~/hw13_microviz.Rmd /xdisk/bhurwitz/bh_class/NETID/assignments/13_microviz
```

# Session info

Records your package versions etc.
Useful for debugging / reproducing analysis.

```{r}
devtools::session_info()
```
